Deciding Bitvector Termination with Program Synthesis

### 2014-07-03 papers/termination/kalashnikov.tex

%\subsection{\newC}
%\label{sec:logic}
\subsection{The Synthesis Specification}
Our (non-)termination specifications are given as program fragments
%More specifically, the logic we use to express the synthesis formula
in a subset of C that we call \newC.  
The characteristic property of a \newC  program is that safety can be decided for
it using a single query to a Bounded Model Checker.  A \newC program is
just a C program with the following syntactic restrictions:
 all loops in the program must have a constant bound;
 all recursion in the program must be limited to a constant depth;
 all arrays must be statically allocated (i.e. not using \texttt{malloc}),
 and be of constant size.
Additionally, \newC programs may use nondeterministic values, assumptions
and arbitrary-width types.

@@

\subsection{The Synthesis Algorithm}
As the quantifier free first-order fragment of the logic is efficiently decidable and
$P$ is expressible as a ground term of the logic, we can use
Counterexample Guided Inductive Synthesis (CEGIS)~\cite{lezama-thesis,sketch} to
check the validity of the synthesis formula.

The core of the CEGIS algorithm is
the refinement loop shown in Fig.~\ref{fig:abstract-refinement} and
detailed in Algorithm~\ref{fig:abstract-refinement-code}.  
The algorithm is divided into two
components: {\sc synth} and {\sc verif}. For a finite set of test
inputs that we track, {\sc synth} synthesises a candidate program that is correct for those tests.
Subsequently, {\sc Verif} verifies the candidate program by searching for an input on
which it does not satisfy the specification. If no such input
can be found, the candidate program is correct.  Otherwise, the new input is added
to the set of tracked inputs.  
%If the input set is finite, this procedure is guaranteed to terminate.

@@

In order to determine the validity of the {\sc synth} formula, we can
check the {\sc synth} program for safety.  The {\sc synth} program is a
\newC program, which means we can check its safety by invoking a Bounded
Model Checker, such as {\sc cbmc}.

In some cases, it is faster to use an explicit-state model checker rather
than a Bounded Model Checker.  This is particularly true when we are checking
the {\sc verif} formula, where we have observed that incorrect programs tend
to be incorrect on a large fraction of the input space.  Counterexamples
are then very easy to find by explicit enumeration of a few inputs.
Since \newC is a fragment of C, we can generate an explicit-state
model checker using the same source files that we pass to {\sc cbmc}
and adding a small function to enumerate possible inputs.
We can run the explicit-state model checker
in parallel with {\sc cbmc} and take the answer of whichever happens
to terminate first, stopping the other process.  This procedure is
depicted for {\sc synth} in Fig.~\ref{fig:synth-dfd}, and is similar for {\sc verif}.

@@

\node [file] (spec) {\sc Specification};
%% \path (synth.south)+(0.0, -0.5) node [file] (tests) {\sc tests.c};
%% \path (tests.south)+(0.0, -0.5) node [file] (interpreter) {\sc interpreter.c};
%% \path (interpreter.south)+(0.0, -0.5) node [file] (spec) {\sc spec.c};

%% \path (spec.east)+(2.0, -0.25) node [process] (merged) {merge};

\path (spec.east)+(2.0, 1.5) node [process] (cbmc) {\sc cbmc};
\path (spec.east)+(2.0, 0.0) node [process] (gcc) {\sc gcc};
\path (spec.east)+(2.0, -1.5) node [process] (gp) {\sc GP};

\path (gcc.east)+(2.5, 0.0) node [file] (out) {Proof};

@@

\path [line] (spec) -- (cbmc);
\path [line] (spec) -- (gcc);
\path [line] (spec) -- (gp);

>>>>

%\subsection{\newC}
%\label{sec:logic}
\subsection{The Synthesis Specification}
The logic we use to express the synthesis formula, i.e. the (non-)termination specifications, % are given as program fragments
%More specifically, the logic we use to express the synthesis formula
is a subset of C that we call \newC.  
The characteristic property of a \newC  program is that safety can be decided for
it using a single query to a Bounded Model Checker.  A \newC program is
just a C program with the following syntactic restrictions:
 all loops in the program must have a constant bound;
 all recursion in the program must be limited to a constant depth;
 all arrays must be statically allocated (i.e. not using \texttt{malloc}),
 and be of constant size.
Additionally, \newC programs may use nondeterministic values, assumptions
and arbitrary-width types.

@@

@@

\subsection{The Synthesis Algorithm}
As safety of a \newC program is efficiently 
%the quantifier free first-order fragment of the logic is efficiently 
decidable and a termination proof is expressible as a ground term of the \newC logic, we use
Counterexample Guided Inductive Synthesis (CEGIS)~\cite{lezama-thesis,sketch} to
check the validity of the synthesis formula. 
The core of the CEGIS algorithm is the refinement loop shown in Fig.~\ref{fig:abstract-refinement} and
detailed in Algorithm~\ref{fig:abstract-refinement-code}.  
The algorithm is divided into two
components: {\sc synth} (see Figure~\ref{fig:synth-dfd}) and {\sc verif}.% described next.

Given a (non-)termination specification (in the synthesis fragment)    
and a finite set of test inputs tracked so far, {\sc synth} generates a program \texttt{synth.c} asserting that some non-deterministic proof taken as input  
is incorrect on at least one of the tracked inputs.
Thus, \texttt{synth.c} is unsafe iff there is some (non-)termination proof that is correct for all of the tracked inputs.  
By using testing techniques, we search for such a proof, i.e. 
we synthesise a candidate (non-)termination proof that is correct for all the tracked inputs.  
The testing techniques we use are: model checking,
exhaustive search and genetic programming (see Section~\ref{sec:gp}).  
%For the latter 2 techniques, we combine SYNTH with another program
%(GP or EXPLICIT respectively) to create a program that searches for test vectors (proofs) that cause SYNTH
%to crash.
As \texttt{synth.c}
is a \newC program, we can check its safety by invoking a Bounded
Model Checker, such as {\sc cbmc} \cite{}.
We run the three available testing techniques %explicit-state model checker
in parallel and take the answer of whichever happens
to terminate first, stopping the other processes.

Subsequently, {\sc Verif} uses model checking or exhaustive search to verify the candidate proof by searching for an input on
which it does not satisfy the specification. If no such input
can be found, the candidate proof is correct.  Otherwise, the new input is added
to the set of tracked inputs and the CEGIS loop is reiterated.

%If the input set is finite, this procedure is guaranteed to terminate.

%% In some cases, it is faster to use an explicit-state model checker rather than a Bounded Model Checker.  
%% This is particularly true for the {\sc verif} component, 
%% where we have observed that incorrect programs tend
%% to be incorrect on a large fraction of the input space.  Counterexamples
%% are then very easy to find by explicit enumeration of a few inputs.
%% Since \newC is a fragment of C, we can generate an explicit-state
%% model checker using the same source files that we pass to {\sc cbmc}
%% and adding a small function to enumerate possible inputs.

%% ---  We begin with the program under analysis (the PUA).  From this we produce a specification (in second order logic)
%% encoding the termination of the PUA.  From the logical specification, we produce a program SYNTH.
%% SYNTH takes as input a proof and asserts that the proof is incorrect on at least one of the inputs being tracked so far.
%% Therefore, SYNTH is unsafe iff there is some proof that is correct for all of the tracked inputs.  Now we use testing
%% techniques to try to find a proof that causes SYNTH to crash.  The testing techniques we use are: model checking,
%% exhaustive search and genetic programming.  For the latter 2 techniques, we combine SYNTH with another program
%% (GP or EXPLICIT respectively) to create a program that searches for test vectors (proofs) that cause SYNTH
%% to crash.

%% In the {\sc synth} component, in order to find a candidate proof $P$ satisfying
%% $\sigma(i_1, P(i_1)) \land \ldots \land \sigma(i_N, P(i_N))$
%% for the tracked inputs $(i_1, \ldots, i_N)$, 
%% we attempt a safety proof.
%In order to determine the validity of the {\sc synth} formula, we can
%check the {\sc synth} program for safety.

@@

\node [file] (spec) {\sc Termination Specification};
\path (spec.south)+(0.0, -1) node [file] (inputs) {\sc Tracked Inputs};
%% \path (synth.south)+(0.0, -0.5) node [file] (tests) {\sc tests.c};
%% \path (tests.south)+(0.0, -0.5) node [file] (interpreter) {\sc interpreter.c};
%% \path (interpreter.south)+(0.0, -0.5) node [file] (spec) {\sc spec.c};

%% \path (spec.east)+(2.0, -0.25) node [process] (merged) {merge};
\path (spec.east)+(2.0, -0.75) node [file] (file) {synth.c};

\path (file.east)+(2.0, 1.5) node [process] (cbmc) {\sc cbmc};
\path (file.east)+(2.0, 0.0) node [process] (gcc) {\sc Search};
\path (file.east)+(2.0, -1.5) node [process] (gp) {\sc GP};

\path (gcc.east)+(2.5, 0.0) node [file] (out) {Candidate Proof};

@@

\path [line] (spec) -- (file);
\path [line] (inputs) -- (file);

\path [line] (file) -- (cbmc);
\path [line] (file) -- (gcc);
\path [line] (file) -- (gp);
